NumPy---Array Programming in Python

### 2020-02-13 body.tex
Add sentence about computational steering
Refine protocols paragraph

Two Python array packages existed before NumPy.
The Numeric package began in the mid-1990s and provided an array object and array-aware functions
in Python, written in C, and linking to standard fast implementations of linear
algebra \cite{dubois1996numerical,Numericmanual}.
% FIXME steering physics codes at llnl
To handle large astronomical images coming from the Hubble Space Telescope,
a reimplementation of Numeric, called Numarray, added
support for structured arrays, flexible indexing, memory mapping, byte-order variants,
more efficient memory use, flexible IEEE error handling capabilities, and
better type casting rules \cite{greenfield2003numarray}.
While Numarray was highly compatible with Numeric, the two packages
had enough differences that it divided the community, until
2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
and Numarray \cite{oliphant2006guide}---combining Numarray's features with Numeric's
performance on small arrays and its rich C \emph{Application Programming Interface} (API).

@@

Ideally, operating on specialized arrays using NumPy functions or semantics would
simply work, so that users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, distributed arrays,
and so forth, as appropriate.
To support array operations between external array objects, NumPy,
therefore, added the capability to act as a central coordination
mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).

@@

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
NumPy, in turn, dispatches operations to the originating library, as required.
Currently around 50 or so functions, which are the most commonly
used functions, support these protocols.
This list is being refined as we gain more experience with how it works in
practice.

Thanks to these developments, users can now, for example, scale
their computation from a single machine to distributed systems using Dask.
The protocols \textit{compose} well, and so it is possible to redeploy NumPy
code onto multi-GPU systems with minimal changes\cite{entschev2019}.

>>>>

Two Python array packages existed before NumPy.
The Numeric package began in the mid-1990s and provided an array object and array-aware functions
in Python, written in C, and linking to standard fast implementations of linear
algebra \cite{dubois1996numerical,Numericmanual}.
One of its earliest uses was to steer C++ applications for
inertial confinement fusion research at
Lawrence Livermore National Laboratory \cite{yang1997steering}.
To handle large astronomical images coming from the Hubble Space Telescope,
a reimplementation of Numeric, called Numarray, added
support for structured arrays, flexible indexing, memory mapping, byte-order variants,
more efficient memory use, flexible IEEE error handling capabilities, and
better type casting rules \cite{greenfield2003numarray}.
While Numarray was highly compatible with Numeric, the two packages
had enough differences that it divided the community, until
2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
and Numarray \cite{oliphant2006guide}---combining Numarray's features with Numeric's
performance on small arrays and its rich C \emph{Application Programming Interface} (API).

@@

Ideally, operating on specialized arrays using NumPy functions or semantics would
simply work, so that users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, distributed arrays,
and so forth, as appropriate.
To support array operations between external array objects, NumPy
therefore added the capability to act as a central coordination
mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).

@@

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for specialized arrays to be
passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
NumPy, in turn, dispatches operations to the originating library, as required.
Over three hundred of the most popular
NumPy functions are supported.  The protocols are implemented by at
least five libraries, including Dask, CuPy, xarray, and PyData/Sparse.
Thanks to these developments, users can now, for example, scale
their computation from a single machine to distributed systems using Dask.
The protocols \textit{compose} well, and so it is possible to redeploy NumPy
code onto multi-GPU systems with minimal changes\cite{entschev2019}.
Working with the community, we have identified several issues that
prevent further adoption, and are refining interfaces to reduce
impedance mismatch.
